---
title: "NYPD Statistics"
author: "L. Himmelspach"
date: "2024-12-02"
output:
  html_document: default
  pdf_document: default
---

I am starting by reading the data in from the url.
Two questions of interest to explore from this data are are there seasonal patterns in shooting incidents, and which of the boroughs have historically had highest rates of shootings?
The data is from https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic 
It is a list of every shooting incident that has occurred in NYC since 2006 through 2023. The data includes an incident key number, the time and date the shooting incident occurred, the borough of the incident, the precinct, victim's age, gender, and race, as well as the coordinates of the shooting.
```{r}
# Install the tidyverse, plotly, dplyr, caret and ggplot2 packages if not already installed
if(!require("tidyverse")){
  install.packages("tidyverse")
}
if(!require("plotly")){
  install.packages("plotly")
}
if(!require("dplyr")){
  install.packages("dplyr")
}
if(!require("ggplot2")){
  install.packages("ggplot2")
}
if(!require("lubridate")){
  install.packages("lubridate")
}
if(!require("forecast")){
  install.packages("forecast")
}

# Load the packages
library(tidyverse)
library(plotly)
library(dplyr)
library(ggplot2)
library(lubridate)
library(forecast)
```

```{r get_jhu_data}
## Get data from csv file
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
```
Start by reading in the data set.

```{r import_data, message=FALSE}
ny_shooting <- read_csv(url_in)
```

I'll be analyzing the number of shootings per date in each borough, so won't need the rest of the data.

```{r tidy_data}
#Tidy the data
df_tidy <- ny_shooting %>% mutate(OCCUR_DATE = mdy(OCCUR_DATE))%>%
  #Convert to date-time format
group_by(OCCUR_DATE,BORO)%>%
#Group by both date and borough
  summarize(total_shootings = n())
  #Count shootings per day
#Explore the data
summary(df_tidy)
```

To analyze the data a bit, we'll look at the shootings over time.

```{r data_analysis}

plot_ly(x = df_tidy$OCCUR_DATE,type="histogram",
        marker = list(color="steelblue")) %>%
  layout(title = "Number of Shooting Incidents Over Time",
         xaxis = list(title = "Date"),
         yaxis = list(title = "Number of Incidents"))

```

```{r model_data}

df_tidy <- df_tidy %>% mutate(OCCUR_DATE = ymd(OCCUR_DATE)) %>% 
  #Convert to date format
  arrange(OCCUR_DATE)
```

```{r time_series_data}
#Create a time series object
ts_shootings <- ts(df_tidy$total_shootings, start = c(min(year(df_tidy$OCCUR_DATE)), 1),end = c(max(year(df_tidy$OCCUR_DATE)), 12), frequency = 12)

#Visualize the Time Series
plot(ts_shootings, main = "Total Shootings Over Time", xlab = "Year", ylab = "Total Shootings")

#Decompose the Time Series
decomposed_ts <- decompose(ts_shootings)

#Plot the decomposed time series
plot(decomposed_ts)

#ARIMA Model
arima_model <- auto.arima(ts_shootings)

#Print the model summary
summary(arima_model)

#Forecast using ARMIA
forecast_arima <- forecast(arima_model, h = 12)

#Plot the ARMIA forecast
plot(forecast_arima, main = "ARMIA Forecast")

#ETS Model
ets_model <- ets(ts_shootings)

#Print the model summary
summary(ets_model)

#Forecast using ETS
forecast_ets <- forecast(ets_model, h = 12)

#Plot the ETS forecast
plot(forecast_ets, main = "ETS Forecast")


```



```{r visualize_data}
#Calculate cumulative shootings
df_tidy <- df_tidy %>% group_by(BORO) %>% mutate(cumulative_shootings = cumsum(total_shootings))
```

```{r cumulative_plot_data}
#Create the cumulative plot
fig <- plot_ly(df_tidy,x=~OCCUR_DATE,y=~cumulative_shootings,color=~BORO,type = 'scatter',mode = 'lines+markers')

#Customize the plot
fig <- fig %>% layout(title = "Cumulative Shootings in NYC by Borough Over Time",xaxis=list(title = "Date"),yaxis = list(title = "Cumulative Shootings"))
#Display the plot
fig
```
Brooklyn has had the most shootings total, and Staten Island has had least
There looks like there is a little jump in activity in 2020
Could be because of COVID, or possibly reaction to George Floyd's murder.

This analysis only includes total shootings, it doesn't include shootings
in terms of events per population.

```{r monthly_data}
#ensure date format is correct
df_tidy$OCCUR_DATE <- as.Date(df_tidy$OCCUR_DATE, format = "%Y-%m-%d")

#Extract month and year
df_tidy <- df_tidy %>%
  mutate(month = 
month(OCCUR_DATE), year = year(OCCUR_DATE))

#Group and summarize
df_summary <- df_tidy %>% 
  group_by(month, year) %>%
  summarize(total_shootings = n())
```

```{r plot_monthly_data}
#Create plot of shootings by month
ggplot(df_summary, aes(x = month, y = total_shootings, fill = factor(year))) +
  geom_bar(stat = "identity") + 
  labs(title = "Shootings by Month and Year",
       x = "Month",
       y = "Total Shootings") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  theme_minimal()
```

It looks like summer months have the highest rates of shootings in NYC.
apparently the nicer weather is not enough to discourage it
February has the least. Possibly because nobody wants to be outside
Also possibly because there are fewer days.

In conclusion, Brooklyn has had the most shootings total, and Staten Island has had least. There were also more shooting incidents on average in the summer months, and fewer in the winter.
Possible sources of bias include reporting discrepancies, under reporting, or classification errors. 

